Binning Strategies and Related Loss for Large Data

### 2013-03-13 Paper.Rnw
Karsten's changes

\documentclass[letterpaper,12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amsthm,amssymb}
\usepackage{fullpage}
\usepackage{graphicx,float,wrapfig,subfig,tabularx,ulem}
\graphicspath{{figure/}}
\usepackage{epstopdf}
%\usepackage{csquotes}
\usepackage{color}
\usepackage{natbib}
\usepackage{hyperref}

%Karsten was here
%Heike, too

@@

%opening
\title{Binning Strategies and Related Loss for Graphics of Large Data}
\author{Susan VanderPlas$^1$, Karsten Maurer$^1$, Heike Hofmann$^{1,2}$\\$^1$Department of Statistics, $^2$Human Computer Interaction\\Iowa State University}

\begin{document}
% \SweaveOpts{concordance=TRUE, warning=FALSE,message=FALSE}
\maketitle

\begin{abstract}
Dealing with the data deluge of the Big Data Age is both exciting and challenging. Large data demands require us to re-think strategies of visualizing data. Binning methods have been suggested in the past as viable alternative to standard plots based on raw data, as the resulting area plots tend to not be affected by increases in data as much. This comes with the price of loss of information inherent to any binning scheme. In this paper we discuss properties of two commonly used binning algorithms. We define loss of information in the specific setting of two dimensional displays, provide readily applicable tools for loss evaluation and discuss the two binning schemes with respect to their loss in the frame work of a simulation and two case studies.
\end{abstract}

\section{Introduction}
<<setup,echo=F,include=F,eval=T>>=
library(ggplot2)
library(gridExtra)
library(dbData)
library(devtools)
library(xtable)
library(RColorBrewer)
opts_chunk$set(concordance=TRUE, tidy=TRUE, cache=TRUE, message=FALSE, warning=FALSE)

@@

  res <- data.frame(sapply(1:(k-1), function(var) {
			return(rep(x[,var], x[,k]))
		}
	))
	names(res) <- names(x)[1:(k-1)]
	return(res)
}
d1.ext <- xtd.data(d1)
@
<<scatterplots, dependson='setup', echo=F, include=FALSE>>=
library(hexbin)
qplot(data=d1.ext, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", alpha=I(.05), size=I(3)) + theme_bw()
  ggsave("./figure/OverplottingAlpha.pdf", width=5, height=5)
qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", size=I(3)) + theme_bw()
  ggsave("./figure/Overplotting.pdf", width=5, height=5)
qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", shape="o", size=I(6)) + scale_shape_identity() + theme_bw()
  ggsave("./figure/OverplottingCircles.pdf", width=5, height=5)
qplot(data=d1.ext, x=G, y=SO, geom="hex", fill=..count.., group=1) +  
  scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 10, 100, 1000)) + theme_bw() + theme(legend.position="bottom")
  ggsave("./figure/HexBinning.pdf", width=5, height=5)
d1.hex.obj <- hexbin(d1.ext$G, d1.ext$SO)
d1.hex <- data.frame(hcell2xy(d1.hex.obj))
names(d1.hex) <- c("G", "SO")
d1.hex$Freq <- d1.hex.obj@count
qplot(data=d1.hex, x=G, y=SO, geom="point", size=Freq) + 
  scale_size_area("Frequency", max_size=4, trans="log", breaks=c(1, 10, 100, 1000, 10000))+ theme_bw() +
  theme(legend.position="bottom")
  ggsave("./figure/BinScatter.pdf", width=5, height=5)

# Sunflower plot
# d1.hex$id <- seq(1:nrow(d1.hex))
# d1.hex2 <-  ddply(d1.hex, .(G, SO, id), function(i) return(data.frame(r = cos((floor(i$Freq/10))*seq(0, 2*pi, .05)), theta=seq(0, 2*pi, 0.05), Freq=i$Freq)))
# ggplot(d1.hex2) + geom_subplot(aes(x=G, y=SO, group=id,  subplot=geom_star(aes(r=r, angle=theta, fill=log(Freq), colour=log(Freq)), r.zero=TRUE)))
@

@@

\citet{Friedman97} proposed in his paper on datamining and statistics that ``Every time the amount of data increases by a factor of ten, we should totally rethink how we analyze it". The same holds for visualizations.  
With a 100-1000 fold increase in the amount of data, the utility of some of  graphs changes dramatically, with some of our most-used tools such as the scatterplot deteriorating quickly \citep{gold}.

% binning as a way out
Area plots, such as histograms, do not tend to be as affected by increases in the amount of data as plots, which utilize  raw data. 
This fact is being exploited to make scatterplots again useful instruments for large data settings  by employing binning strategies \citep{gold}.

@@

\section{Scatterplots for Large Data Sets}
In the case of medium sized data scatterplots are great tools for showing relationships in two dimensions. For large data, scatterplots suffer from over-plotting -- i.e. more and more points are drawn in close-by places, thereby masking relevant structure.
Figure~\ref{fig:scatter-alpha} shows an example taken from baseball statistics. The scatterplot shows 139 seasons (from the years of 1871 -- 2009) of  pitching statistics for every baseball players as published in Sean Lahman's Baseball database (\url{http://www.seanlahman.com/baseball-archive/}).  The number of games (G) played in a season is plotted versus number of strikeouts (SO), i.e. a batter receives three strikes during his time at bat. Pitchers want to throw as many strikeouts as possible. While the data set is only medium sized with \Sexpr{dim(d1.ext)[1]} observations, it already shows some of the break-down patterns scatterplots experience with large data.
Figure~\ref{fig:scatter-alpha} shows a traditional scatterplot on the left. Each observation is drawn with a filled circle. A triangular structure is apparent with some outliers at a medium number of games and high number of strikeouts. Tukey \citep{tukey} suggested the use of open circles (see Figure~\ref{fig:scatter-alpha}b) to mitigate the problem of over-plotting. Open circles make points visible that are close together. This technique is  not suitable to determine the magnitude, but gives more information than is available with filled points. 
A modern alternative to open circles is alpha-blending (see Figure~\ref{fig:scatter-alpha}c). Alpha blending provides  more frequency information, allowing a distinction of higher-resolution density.

\begin{figure}[hbtp]
  \subfloat[Overplotted data ]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{figure/Overplotting.pdf}}
  \subfloat[Tukey-style open circles]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{figure/OverplottingCircles.pdf}} 
	\subfloat[Alpha blending]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{figure/OverplottingAlpha.pdf}}

  \subfloat[Hexagonal binning]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{figure/HexBinning.pdf}}
  \subfloat[Scatterplot of Binned Data]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{figure/BinScatter.pdf}}

@@

Histograms are an example of plots which use binned variables. To extend histograms to a graph which can represent the joint distribution between two variables, it is a natural step to create a two-dimensional graph which uses some other attribute (color, 3D rendering) to provide joint density information. A {\it tile plot} uses shading to provide frequency information, with tiles (rather than bars in a histogram) centered at the bin center, rather like a two-dimensional histogram viewed from above.

@@

When the point area is scaled in proportion to frequency, scatterplots can be used to display the binned data as well.  When points are filled circles, these plots are also known as ``bubble plots", which were first used by William Playfair \citep{playfair, playfair2}. Kernel density smoothing can be used to vary $\alpha$ or color according to a smoothed density, providing features similar to tile plots or $\alpha$-blended scatterplots in a more smooth, continuous fashion. These estimates require parameter tuning and may hide gaps in the data by over-smoothing while simultaneously de-emphasizing outlying points.

As alternatives, Figure~\ref{fig:scatter-alpha} (d-e) contains examples of a hexagonal tile plot with frequency encoded as color (d) and a ``bubble plot" (referred to as a modified scatterplot) with frequency encoded as point size (e). The hexagonal tiles and the modified scatterplot are more effective at displaying the shape of the joint density and preserving outliers than any of the scatterplots shown in Figure~\ref{fig:scatter-alpha} (a-c). The tile plot is less prone to the Hermann-grid illusion than the bubble plot, particularly for binned data, where bin centers usually fall on a grid.

% \begin{figure}[hbtp]
% <<conclusions, out.width='\\textwidth', fig.width=5, fig.height=3>>=
% library(dbData)
% connect <- dbConnect(dbDriver("MySQL"), user="2009Expo",
% password="R R0cks", port=3306, dbname="baseball",
% host="headnode.stat.iastate.edu")
% pitch <- new("dataDB", co=connect, table="Pitching")
% pitch.stats <- dbData(vars=list("G", "SO", "yearID"), pitch)
% qplot(yearID, SO/G, alpha=I(0.2), weight=Freq, data=pitch.stats, colour=G) + xlab("season")
% @
% \caption{\caption{strikeout-rate}Scatterplot of rate of strikeouts per game over time. Each dot corresponds to a pitcher's seasonal average. Coloring indicates number of games played, alpha-blending is used additionally to emphasize high-density regions. There's clear break points in the strikeout rates over time. In 1902 XXXX}
% \end{figure}
Only in the tile and the bubble plot the inner structure of the data becomes apparent: 
the joint density consists two distinct ridges following two lines with very different slopes. The lower slope corresponds to the --modern-- average strike out rate of pitchers of just under one strike-out per game. The other line has a slope of about four times that rate. This high rate is also associated with fewer games played. Closer investigation of other, related variables reveals that this high strike-out rate corresponds mainly to historic pitchers with much shorter seasons (in 1876 only 70 games were played in a season, as opposed to 162 in 2009), and qualitatively different balls and  bats.

@@

For extremely large data sets, tile plots are a more useful visualization of two-dimensional density information than the scatterplot, and are less computationally demanding, as not every single point in the data set has to be  rendered separately.

@@

Figure~\ref{fig:BinningTutorial} provides an illustration of the binning process.
<<binningDemo, echo=FALSE, include=FALSE>>=
      library(mvtnorm)
      set.seed(33)
      d <- rmvnorm(200, c(5, 25), matrix(c(100, 50, 50, 100), nrow=2))
      d <- data.frame(d)
      names(d) <- c("x", "y")
      d <- subset(d, (x>-20 & x<40 & y>0 & y<50),drop=TRUE)
      d.rand <- d
      dnew <- binStd(cbind(d-5, Freq=1), c(10, 10))
      dnew$x <- dnew$x+5
      dnew$y <- dnew$y+5
      d$bin <- dnew$x%in%c(5,15) & dnew$y==25
      d$a <- .33 + .66*d$bin
      d.orig <- d

@@

      temp <- data.frame(   
                x=c(-20, -10,   0,  10,  20,  30,  40, -20, -20, -20, -20, -20, -20), 
             xend=c(-20, -10,   0,  10,  20,  30,  40,  40,  40,  40,  40,  40,  40), 
                y=c(  0,   0,   0,   0,   0,   0,   0,   0,  10,  20,  30,  40,  50), 
             yend=c( 50,  50,  50,  50,  50,  50,  50,   0,  10,  20,  30,  40,  50), 
            group=c(  1,   2,   3,   4,   5,   6,   7,   8,   9,  10,  11,  12,  13))
      qplot(data=d[,], x=x, y=y, alpha=a,  geom="point", 
            xlim=c(-20, 40), ylim=c(0, 50)) +
        annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), 
                 fill="blue", alpha=.25) +
        annotate("polygon", x=c(0, 0, 10, 10), y=c(20, 30, 30, 20), 
                 fill="green", alpha=.25) +
        scale_alpha(range=c(.5, 1), guide=FALSE) +
        scale_colour_identity(guide=FALSE) + 
        scale_shape(solid=T,guide=FALSE) +
        theme_bw() +
        annotate("point", x=rep(seq(-15, 35, 10),each=7), 
               y=rep(seq(-5, 55, 10), times=6), 
               shape='+', size=8, colour="gray90") + 
        geom_point(data=d[which(!is.na(d$xbin)),], 
                   aes(x=x, y=y, shape=factor(xbin), colour=factor(xbin)), 
                   size=I(3), aes.inherit=FALSE) + 
        scale_shape_manual(name="Bin", values=c("5"=16, "15"=15), 
                           labels=c("(5,25)", "(15, 25)"),guide=FALSE) +
        scale_colour_manual(values=c("5"="green", "15"="blue"), name="Bin", 
                            labels=c("(5,25)", "(15, 25)"),guide=FALSE)+
        annotate("segment", x=temp$x, y=temp$y, 
                 xend=temp$xend, yend=temp$yend, group=temp$group)  + ylim(c(0, 50))
      #ggsave("./figure/Binning-DataAndBin.pdf", width=5, height=5, units="in")

      d.orig$bin <- dnew$x==15 & dnew$y==25
      d.orig$a <- .33+.66*d.orig$bin
      d.orig$l <- "Data outside bin"
      d.orig$l[d.orig$bin] <- "Binned Data"
      d.orig$size <- 2
      d.orig$size[d.orig$bin] <- 3
      d.orig <- rbind(d.orig, data.frame(x=15, y=25, a=1, l="Visual Center", size=2,bin=TRUE), 
                    data.frame(x=mean(d.orig$x[d.orig$bin]), y=mean(d.orig$y[d.orig$bin]), 
                               l="Numerical Center", a=1, size=2, bin=TRUE))
      ggplot(data=d.orig, aes(x=x, y=y, shape=l, size=l, colour=l)) + 
        annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), 
                 fill="blue", alpha=.10) + 
        geom_point() +
        scale_shape_manual(name="", values=c("Data outside bin" = 16, 
                                             "Binned Data" = 15, 
                                             "Numerical Center" = 17, 
                                             "Visual Center" = 18)) +
        scale_colour_manual(name="", values=c("Data outside bin" = "grey60", 
                                              "Binned Data" = "blue",
                                              "Numerical Center" = "blue4", 
                                              "Visual Center" = "blue4"))+
        scale_size_manual(name="", values=c("Data outside bin" = 3, "Binned Data"=3,
                                            "Numerical Center" = 5, 
                                            "Visual Center" = 5))+
        theme_bw() +
        annotate("point", x=15, y=25, shape=18, size=5, colour="blue4") + 
        annotate("point", x=d.orig$x[which(d.orig$l=="Numerical Center")], 
                 y=d.orig$y[which(d.orig$l=="Numerical Center")], shape=17, size=5, 
                 colour="blue4") +
        annotate("segment", x=temp$x, y=temp$y, xend=temp$xend, yend=temp$yend, group=temp$group) + 
#         annotate("text", x=d$x[which(d$l=="Numerical Center")], 
#                  y=d$y[which(d$l=="Numerical Center")], 
#                  label=expression(paste('(', over(phantom(0),x)[i], ',', over(phantom(0),y)[i],')'))) +
        scale_x_continuous(breaks=c(10, 15, 20), limits=c(8, 22), labels=c(expression(x^{"*"}-b[x]/2), expression(x^{"*"}), expression(x^{"*"}+b[x]/2)), name='') + 
        scale_y_continuous(breaks=c(20, 25, 30), limits=c(18, 32), labels=c(expression(y^{"*"}-b[y]/2), expression(y^{"*"}), expression(y^{"*"}+b[y]/2)), name='')
      #ggsave("./figure/Binning-DataVisualNumericalCenter.pdf", width=5.5, height=3.25, units="in")

@@

\begin{figure}[hbtp]
  \centering
  \subfloat[Standard Algorithm]{\includegraphics[keepaspectratio=true, width=3in]{./figure/binningDemo1.pdf}}
%   \subfloat{\includegraphics[keepaspectratio=TRUE, width=2in]
%   {./figure/RandomBinningDataPoints1}}
  \subfloat[Random Algorithm]{\includegraphics[keepaspectratio=TRUE, width=3in]
  {./figure/RandomBinDemo.pdf}}
  \caption{Points are assigned to bin (0, 20) and bin (10,20). Points in the standard algorithm are all contained within the bin region, where points in the random algorithm may be outside the direct region.}
  \label{fig:BinningTutorial}
\end{figure}

\subsection{Random Binning Algorithm}
<<RandomBinDemo,echo=F,include=FALSE,cache=TRUE>>=
nl <- ncol(d.rand)
dnew <- d.rand
binning <- c(10, 10)
for(x in 1:(nl)){
  y <- d.rand[,x] - binning[x]/2
dnew[,x] <- binning[x]*(floor(y/binning[x]) + 
  sapply((y%%binning[x])/binning[x], function(p) rbinom(1, 1, p))) + binning[x]/2
}
temp <- data.frame(   
          x=c(-20, -10,   0,  10,  20,  30,  40, -20, -20, -20, -20, -20, -20), 
       xend=c(-20, -10,   0,  10,  20,  30,  40,  40,  40,  40,  40,  40,  40), 
          y=c(  0,   0,   0,   0,   0,   0,   0,   0,  10,  20,  30,  40,  50), 
       yend=c( 50,  50,  50,  50,  50,  50,  50,   0,  10,  20,  30,  40,  50), 
      group=c(  1,   2,   3,   4,   5,   6,   7,   8,   9,  10,  11,  12,  13))

>>>>

\documentclass[letterpaper,12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amsthm,amssymb}
\usepackage{fullpage}
\usepackage{graphicx,float,wrapfig,subfig,tabularx,ulem}
\graphicspath{{images/}}
\usepackage{epstopdf}
%\usepackage{csquotes}
\usepackage{color}
\usepackage{natbib}
\usepackage{hyperref}

@@

%opening
\title{Binning Strategies and Related Loss for Large Data}
\author{Susan VanderPlas$^1$, Heike Hofmann$^{1,2}$\\$^1$Department of Statistics, $^2$Human Computer Interaction\\Iowa State University}

\begin{document}
\SweaveOpts{concordance=TRUE, warning=FALSE,message=FALSE}
\maketitle

\begin{abstract}
Dealing with the data deluge of the Big Data Age is both exciting and challenging. The demands of large data require us to re-think strategies of visualizing data. Plots employing binning methods have been suggested in the past as viable alternative to standard plots based on raw data, as the resulting area plots tend to not be affected by increases in data as much. This comes with the price of loss of information inherent to any binning scheme. In this paper we discuss properties of two commonly used binning algorithms. We define loss of information in the specific setting of two dimensional displays, provide readily applicable tools for loss evaluation and discuss the two binning schemes with respect to their loss in the framework of a simulation and two case studies.
\end{abstract}

\section{Introduction}
<<setup,echo=F,include=F,eval=T>>=
library(ggplot2)
library(gridExtra)
library(dbData)
#library(multicore)
library(devtools)
library(xtable)
library(RColorBrewer)
# setwd("/home/susan/Dropbox/GraphicsGroup/")
setwd("/Users/heike/Dropbox/papers/")

setwd("./dbaccess/NewWriteUp/")

@@

  res <- data.frame(sapply(1:(k-1), function(var) {
			return(rep(x[,var], x[,k]))
		}
	))
	names(res) <- names(x)[1:(k-1)]
	return(res)
}
d1.ext <- xtd.data(d1)
@
<<scatterplots, echo=F,include=F,eval=F>>=
qplot(data=d1.ext, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", alpha=I(.05), size=I(3)) + theme_bw()
  ggsave("./images/OverplottingAlpha.pdf", width=5, height=5)
qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", size=I(3)) + theme_bw()
  ggsave("./images/Overplotting.pdf", width=5, height=5)
qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", shape="o", size=I(6)) + scale_shape_identity() + theme_bw()
  ggsave("./images/OverplottingCircles.pdf", width=5, height=5)
qplot(data=d1.ext, x=G, y=SO, geom="hex", fill=..count.., group=1) +  
  scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 10, 100, 1000)) + theme_bw() + theme(legend.position="bottom")
  ggsave("./images/HexBinning.pdf", width=5, height=5)
d1.hex.obj <- hexbin(d1.ext$G, d1.ext$SO)
d1.hex <- data.frame(hcell2xy(d1.hex.obj))
names(d1.hex) <- c("G", "SO")
d1.hex$Freq <- d1.hex.obj@count
qplot(data=d1.hex, x=G, y=SO, geom="point", size=Freq) + 
  scale_area("Frequency", range=c(1, 4), trans="log", breaks=c(1, 10, 100, 1000, 10000))+ theme_bw() +
  theme(legend.position="bottom")
  ggsave("./images/BinScatter.pdf", width=5, height=5)

# Sunflower plot
# d1.hex$id <- seq(1:nrow(d1.hex))
# d1.hex2 <-  ddply(d1.hex, .(G, SO, id), function(i) return(data.frame(r = cos((floor(i$Freq/10))*seq(0, 2*pi, .05)), theta=seq(0, 2*pi, 0.05), Freq=i$Freq)))
# ggplot(d1.hex2) + geom_subplot(aes(x=G, y=SO, group=id,  subplot=geom_star(aes(r=r, angle=theta, fill=log(Freq), colour=log(Freq)), r.zero=TRUE)))

@@

\citet{Friedman97} proposed in his paper on data mining and statistics that ``Every time the amount of data increases by a factor of ten, we should totally rethink how we analyze it". The same holds for visualizations.  
With a 100-1000 fold increase in the amount of data, the utility of some of our most used graphical tools, such as scatterplots, deteriorates quickly \citep{gold}.

% binning as a way out
Area plots, such as histograms, do not tend to be as affected by increases in the amount of data as plots that utilize raw data. 
By using binning strategies and the principles for displaying information in area plots, scatterplots can again become useful instruments for large data settings \citep{gold}.

@@

\section{Scatterplots for Large Data Sets}
In the case of medium sized data scatterplots are great tools for showing relationships in two dimensions. For large data, scatterplots suffer from over-plotting -- i.e. more and more points are drawn in close-by places, thereby masking relevant structure.
Figure~\ref{fig:scatter-alpha} shows an example taken from baseball statistics. The scatterplot shows 139 seasons (from the years of 1871 -- 2009) of  pitching statistics for every baseball pitcher as published in Sean Lahman's Baseball database (\url{http://www.seanlahman.com/baseball-archive/}).  The number of games played in a season is plotted versus number of strikeouts a pitchr threw over the course of a season. While the data set is only medium sized with \Sexpr{dim(d1.ext)[1]} observations, it already shows some of the break-down patterns scatterplots experience with large data.
Figure~\ref{fig:scatter-alpha} shows a traditional scatterplot on the left. Each observation is drawn with a filled circle. A triangular structure is apparent with some outliers at a medium number of games and high number of strikeouts. Tukey \citep{tukey} suggested the use of open circles (see Figure~\ref{fig:scatter-alpha}b) to mitigate the problem of over-plotting. Open circles make points visible that are close together. This technique is  not suitable to determine the magnitude, but gives more information than is available with filled points. 
A modern alternative to open circles is alpha-blending (see Figure~\ref{fig:scatter-alpha}c). Alpha blending provides  more frequency information, allowing a distinction of higher-resolution density.

\begin{figure}[hbtp]
  \subfloat[Overplotted data ]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{images/Overplotting.eps}}
  \subfloat[Tukey-style open circles]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{images/OverplottingCircles.eps}} 
	\subfloat[Alpha blending]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{images/OverplottingAlpha.eps}}

  \subfloat[Hexagonal binning]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{images/HexBinning.eps}}
  \subfloat[Scatterplot of Binned Data]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{images/BinScatter.eps}}

@@

Histograms are an example of plots which use binned variables. To extend histograms to a graph which can represent the joint distribution between two variables, it is a natural step to form a tesselated grid on a two dimensional Cartesian plane and use some other attribute (color, 3D rendering) to provide joint density information within each grid cell, known as a tile. A {\it binned scatterplot} uses shading to provide frequency information, with tiles (rather than bars in a histogram) centered at the bin center, rather like a two-dimensional histogram viewed from above.

@@

When the point area is scaled in proportion to frequency, scatterplots can be used to display the binned data as well.  When points are filled circles, these plots are also known as ``bubble plots", which were first used by William Playfair \citep{playfair, playfair2}. Kernel density smoothing can be used to vary $\alpha$ or color according to a smoothed density, providing features similar to binned scatterplots or $\alpha$-blended scatterplots in a more smooth, continuous fashion. These estimates require parameter tuning and may hide gaps in the data by over-smoothing while simultaneously de-emphasizing outlying points.

As alternatives, Figure~\ref{fig:scatter-alpha} (d-e) contains examples of a hexagonally binned scatterplot with frequency encoded as color (d) and a ``bubble plot" (referred to as a modified scatterplot) with frequency encoded as point size (e). The hexagonal tiles and the modified scatterplot are more effective at displaying the shape of the joint density and preserving outliers than any of the scatterplots shown in Figure~\ref{fig:scatter-alpha} (a-c). The binned scatterplot is less prone to the Hermann-grid illusion than the bubble plot, particularly for binned data, where bin centers usually fall on a grid.

% \begin{figure}[hbtp]
% <<conclusions, out.width='\\textwidth', fig.width=5, fig.height=3>>=
% library(dbData)
% connect <- dbConnect(dbDriver("MySQL"), user="2009Expo",
% password="R R0cks", port=3306, dbname="baseball",
% host="headnode.stat.iastate.edu")
% pitch <- new("dataDB", co=connect, table="Pitching")
% pitch.stats <- dbData(vars=list("G", "SO", "yearID"), pitch)
% qplot(yearID, SO/G, alpha=I(0.2), weight=Freq, data=pitch.stats, colour=G) + xlab("season")
% @
% \caption{\caption{strikeout-rate}Scatterplot of rate of strikeouts per game over time. Each dot corresponds to a pitcher's seasonal average. Coloring indicates number of games played, alpha-blending is used additionally to emphasize high-density regions. There's clear break points in the strikeout rates over time. In 1902 XXXX}
% \end{figure}
Only in the binned scatterplot and the bubble plot the inner structure of the data becomes apparent: 
the joint density consists two distinct ridges following two lines with very different slopes. The lower slope corresponds to the modern average strike out rate of pitchers of just under one strike-out per game. The other line has a slope of about four times that rate. This high rate is also associated with fewer games played. Closer investigation of other, related variables reveals that this high strike-out rate corresponds mainly to historic pitchers with much shorter seasons (in 1876 only 70 games were played in a season, as opposed to 162 in 2009), and qualitatively different balls and  bats.

@@

For extremely large data sets, binned scatterplots are a more useful visualization of two-dimensional density information than the scatterplot, and are less computationally demanding, as not every single point in the data set has to be  rendered separately.

@@

Figure~\ref{fig:BinningTutorial} provides an illustration of the binning process.
<<binningDemo,echo=F,eval=F,include=F,cache=T>>=
      library(mvtnorm)
      set.seed(33)
      d <- rmvnorm(200, c(5, 25), matrix(c(100, 50, 50, 100), nrow=2))
      d <- data.frame(d)
      names(d) <- c("x", "y")
      d <- subset(d, (x>-20 & x<40 & y>0 & y<50),drop=TRUE)
      d.rand <- d
      dnew <- binStd(cbind(d-5, Freq=1), c(10, 10))
      dnew$x <- dnew$x+5
      dnew$y <- dnew$y+5
      d$bin <- dnew$x%in%c(5,15) & dnew$y==25
      d$a <- .33 + .66*d$bin
      d.orig <- d

@@

      temp <- data.frame(   
                x=c(-20, -10,   0,  10,  20,  30,  40, -20, -20, -20, -20, -20, -20), 
             xend=c(-20, -10,   0,  10,  20,  30,  40,  40,  40,  40,  40,  40,  40), 
                y=c(  0,   0,   0,   0,   0,   0,   0,   0,  10,  20,  30,  40,  50), 
             yend=c( 50,  50,  50,  50,  50,  50,  50,   0,  10,  20,  30,  40,  50), 
            group=c(  1,   2,   3,   4,   5,   6,   7,   8,   9,  10,  11,  12,  13))
      qplot(data=d[,], x=x, y=y, alpha=a,  geom="point", 
            xlim=c(-20, 40), ylim=c(0, 50)) +
        annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), 
                 fill="blue", alpha=.25) +
        annotate("polygon", x=c(0, 0, 10, 10), y=c(20, 30, 30, 20), 
                 fill="green", alpha=.25) +
        scale_alpha(range=c(.5, 1), guide=FALSE) +
        scale_colour_identity(guide=FALSE) + 
        scale_shape(solid=T,guide=FALSE) +
        theme_bw() +
        annotate("point", x=rep(seq(-15, 35, 10),each=7), 
               y=rep(seq(-5, 55, 10), times=6), 
               shape='+', size=8, colour="gray90") + 
        geom_point(data=d[which(!is.na(d$xbin)),], 
                   aes(x=x, y=y, shape=factor(xbin), colour=factor(xbin)), 
                   size=I(3), aes.inherit=FALSE) + 
        scale_shape_manual(name="Bin", values=c("5"=16, "15"=15), 
                           labels=c("(5,25)", "(15, 25)"),guide=FALSE) +
        scale_colour_manual(values=c("5"="green", "15"="blue"), name="Bin", 
                            labels=c("(5,25)", "(15, 25)"),guide=FALSE)+
        annotate("segment", x=temp$x, y=temp$y, 
                 xend=temp$xend, yend=temp$yend, group=temp$group)  + ylim(c(0, 50))
      ggsave("./images/Binning-DataAndBin.pdf", width=5, height=5, units="in")

      d.orig$bin <- dnew$x==15 & dnew$y==25
      d.orig$a <- .33+.66*d.orig$bin
      d.orig$l <- "Data outside bin"
      d.orig$l[d.orig$bin] <- "Binned Data"
      d.orig$size <- 2
      d.orig$size[d.orig$bin] <- 3
      d.orig <- rbind(d.orig, data.frame(x=15, y=25, a=1, l="Visual Center", size=2,bin=TRUE), 
                    data.frame(x=mean(d.orig$x[d.orig$bin]), y=mean(d.orig$y[d.orig$bin]), 
                               l="Numerical Center", a=1, size=2, bin=TRUE))
      ggplot(data=d.orig, aes(x=x, y=y, shape=l, size=l, colour=l)) + 
        annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), 
                 fill="blue", alpha=.10) + 
        geom_point() +
        scale_shape_manual(name="", values=c("Data outside bin" = 16, 
                                             "Binned Data" = 15, 
                                             "Numerical Center" = 17, 
                                             "Visual Center" = 18)) +
        scale_colour_manual(name="", values=c("Data outside bin" = "grey60", 
                                              "Binned Data" = "blue",
                                              "Numerical Center" = "blue4", 
                                              "Visual Center" = "blue4"))+
        scale_size_manual(name="", values=c("Data outside bin" = 3, "Binned Data"=3,
                                            "Numerical Center" = 5, 
                                            "Visual Center" = 5))+
        theme_bw() +
        annotate("point", x=15, y=25, shape=18, size=5, colour="blue4") + 
        annotate("point", x=d.orig$x[which(d.orig$l=="Numerical Center")], 
                 y=d.orig$y[which(d.orig$l=="Numerical Center")], shape=17, size=5, 
                 colour="blue4") +
        annotate("segment", x=temp$x, y=temp$y, xend=temp$xend, yend=temp$yend, group=temp$group) + 
#         annotate("text", x=d$x[which(d$l=="Numerical Center")], 
#                  y=d$y[which(d$l=="Numerical Center")], 
#                  label=expression(paste('(', over(phantom(0),x)[i], ',', over(phantom(0),y)[i],')'))) +
        scale_x_continuous(breaks=c(10, 15, 20), limits=c(8, 22), labels=c(expression(x^{"*"}-b[x]/2), expression(x^{"*"}), expression(x^{"*"}+b[x]/2)), name='') + 
        scale_y_continuous(breaks=c(20, 25, 30), limits=c(18, 32), labels=c(expression(y^{"*"}-b[y]/2), expression(y^{"*"}), expression(y^{"*"}+b[y]/2)), name='')
      ggsave("./images/Binning-DataVisualNumericalCenter.pdf", width=5.5, height=3.25, units="in")
@

@@

\begin{figure}[hbtp]
  \centering
  \subfloat[Standard Algorithm]{\includegraphics[keepaspectratio=true, width=3in]{./images/Binning-DataAndBin.eps}}
%   \subfloat{\includegraphics[keepaspectratio=TRUE, width=2in]
%   {./images/RandomBinningDataPoints1}}
  \subfloat[Random Algorithm]{\includegraphics[keepaspectratio=TRUE, width=3in]
  {./images/RandomBinningDataPoints2.eps}}
  \caption{Points are assigned to bins centered at (5, 25) an (15,25). Points in the standard algorithm are all contained within the bin region, where points in the random algorithm may be outside the direct region.}
  \label{fig:BinningTutorial}
\end{figure}

\subsection{Random Binning Algorithm}
<<RandomBinDemo,echo=F,include=FALSE,eval=F>>=
nl <- ncol(d.rand)
dnew <- d.rand
binning <- c(10, 10)
for(x in 1:(nl)){
  y <- d.rand[,x] - binning[x]/2
dnew[,x] <- binning[x]*(floor(y/binning[x]) + 
  sapply((y%%binning[x])/binning[x], function(p) rbinom(1, 1, p))) + binning[x]/2
}
temp <- data.frame(   
          x=c(-20, -10,   0,  10,  20,  30,  40, -20, -20, -20, -20, -20, -20), 
       xend=c(-20, -10,   0,  10,  20,  30,  40,  40,  40,  40,  40,  40,  40), 
          y=c(  0,   0,   0,   0,   0,   0,   0,   0,  10,  20,  30,  40,  50), 
       yend=c( 50,  50,  50,  50,  50,  50,  50,   0,  10,  20,  30,  40,  50), 
      group=c(  1,   2,   3,   4,   5,   6,   7,   8,   9,  10,  11,  12,  13))
